---
title: "Machine Learning Block1 Lab2"
author: "Lepeng Zhang, Xuan Wang, Priyarani Patil"
date: "2023-11-24"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The group report was made based on the discussion after all of us had finished all three assignments. 

Assignment 1 was mainly contributed by Priyarani Patil & Lepeng Zhang. 

Assignment 2 was mainly contributed by Xuan Wang & Lepeng Zhang. 

# Assignment 1. Explicit regularization
### Q1
See appendix.
```{r, er1, echo = FALSE}

```


### Q2
See appendix.
```{r, er2, echo = FALSE}

```

### Q3
See appendix.
```{r, er3, echo = FALSE}

```

### Q4
See appendix.
```{r, er4, echo = FALSE}

```


# Assignment 2. Decision trees and logistic regression for bank marketing

### Q1
See appendix.
```{r, tree1, echo = FALSE}

```

# Assignment 3. Principal components and implicit regularization

### Q1
See appendix.
```{r, pca1, echo = FALSE}

```


# Appendix
## Code for Assignment 1
### Q1
```{r ref.label=c('er1'), echo=TRUE, eval=FALSE}

```

### Q2
```{r ref.label=c('er2'), echo=TRUE, eval=FALSE}

```


### Q3
```{r ref.label=c('er3'), echo=TRUE, eval=FALSE}

```


### Q4
```{r ref.label=c('er4'), echo=TRUE, eval=FALSE}

```


## Code for Assignment 2
### Q1
```{r ref.label=c('tree1'), echo=TRUE, eval=FALSE}

```

